---
title: "RcppBessel: Bessel Headers and use with Rcpp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RcppBessel: Bessel Headers and use with Rcpp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `RcppBessel` package exports headers for the Bessel functions found in the 
`Bessel` package of Maechler by wrapping the functions in the C code (translated
and cleaned up from Amos' original FORTRAN code by Maechler) into C++. Not all 
functionality is exposed, in order to keep things simple, but may be expanded 
in the future to export more of the functions such as those for the asymptotic 
expansion of BesselK and BesselI for large nu and x. The package also exports 
the functions for use in R, as it was found to offer a speed up of up to 10x.

The initial motivation for writing this package was so as to allow the modified
Bessel function of the second kind to be used in C++ code which the author needed
in another package. The [Boost](https://www.boost.org/doc/libs/1_85_0/libs/math/doc/html/math_toolkit/bessel/mbessel.html) 
implementation does not allow for complex arguments
whilst the special math functions in [std library](https://en.cppreference.com/w/cpp/numeric/special_functions/cyl_bessel_k)
is not available for Mac OS (via the shipped libc++).  Additionally, neither of those 
implementation has the option of exponential scaling.

# Implementation

The table below shows the functions currently available.


| **RcppBessel Function** | **Description** |
|-------------------------|-----------------|
| `bessel_k`               | Computes the modified Bessel function of the second kind. |
| `bessel_y`               | Computes the Bessel function of the second kind (Neumann function). |
| `bessel_h`               | Computes the Hankel function (Bessel function of the third kind). |
| `bessel_i`               | Computes the modified Bessel function of the first kind.  |
| `bessel_j`               | Computes the Bessel function of the first kind. |
| `airy_a`                 | Computes the Airy function of the first kind. |
| `airy_b`                 | Computes the Airy function of the second kind. |

To distinguish between the CamelCase naming in the `Bessel` package and the lowercamelCase naming in base R,
the package has adopted the snake_case syntax similar to the `std` lib. Other than that, the arguments are
the same as in the Bessel package with the exception of `nSeq`, which computes the result for a whole sequence of nu values, 
which we have opted to skip in the implementation.

For real values less than zero, `bessel_k` and `bessel_y` return complex values. 
In this case, if the vector `z` has negative values, the whole output returned is complex valued. The
table below shows the general input/output types expected.


| Function Name | Input for `z`                        | Output Type  |
|---------------|--------------------------------------|--------------|
| `bessel_k`    | Real (all >= 0)                      | Real         |
|               | Real (any < 0)                       | Complex      |
|               | Complex                              | Complex      |
| `bessel_y`    | Real (all >= 0)                      | Real         |
|               | Real (some < 0)                      | Complex      |
|               | Complex                              | Complex      |
| `bessel_h`    | Real (any value)                     | Complex      |
|               | Complex                              | Complex      |
| `bessel_i`    | Real (any value)                     | Real         |
|               | Complex                              | Complex      |
| `bessel_j`    | Real (any value)                     | Real         |
|               | Complex                              | Complex      |
| `airy_a`      | Real (any value)                     | Real         |
|               | Complex                              | Complex      |
| `airy_b`      | Real (any value)                     | Real         |
|               | Complex                              | Complex      |

For values of zero, the output is similar to what the `Bessel` package returns 
with the exception of `bessel_k` which returns `Inf` similar to base R, whereas
`Bessel::BesselK` returns `0`.


The next sections provide a individual summary of each of the functions and a sample
`C++` code which can be pasted and sourced to test the results against the `Bessel` 
function implementation and also show how the package can be used inside other Rcpp 
functions.

## BesselK (`bessel_k`)

### Description

Computes the modified Bessel function of the second kind.

### Inputs

- `z`: A numeric or complex vector of points at which to evaluate the Bessel function.
- `nu`: A numeric value representing the order of the Bessel function.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A either a numeric or complex vector (depending on the input) with the values 
of the `bessel_k` function.


### Example

```cpp
// test_bessel_k.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_k(SEXP x, double nu, bool expon_scaled, int verbose) {
  SEXP z = RcppBessel::bessel_k(x, nu, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)
# Source the C++ code
Rcpp::sourceCpp("test_bessel_k.cpp")

# Generate a sequence
x <- seq(1, 10, by = 1)

# Compare results with Bessel package
tmp <- Bessel::BesselK(x, 2, FALSE, 1, 0)
all.equal(tmp, test_bessel_k(x, 2, FALSE, 0))
*/
```


## BesselI (`bessel_i`)

### Description

Computes the modified Bessel function of the first kind.

### Inputs

- `z`: A numeric or complex vector of points at which to evaluate the Bessel function.
- `nu`: A numeric value representing the order of the Bessel function.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A numeric or complex vector (depending on the input) with the values of the `bessel_i` function.

### Example

```cpp
// test_bessel_i.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_i(SEXP x, double nu, bool expon_scaled, int verbose) {
  SEXP z = RcppBessel::bessel_i(x, nu, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)
# Source the C++ code
Rcpp::sourceCpp("test_bessel_i.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 1)

# Compare results with Bessel package
tmp <- Bessel::BesselI(x, 2, FALSE, 1, 0)
all.equal(tmp, test_bessel_i(x, 2, FALSE, 0))
*/
```


## BesselJ (`bessel_j`)

### Description

Computes the Bessel function of the first kind.

### Inputs

- `z`: A numeric or complex vector of points at which to evaluate the Bessel function.
- `nu`: A numeric value representing the order of the Bessel function.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A numeric or complex vector (depending on the input) with the values of the `bessel_j` function.

### Example

```cpp
// test_bessel_j.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_j(SEXP x, double nu, bool expon_scaled, int verbose) {
  SEXP z = RcppBessel::bessel_j(x, nu, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)
# Source the C++ code
Rcpp::sourceCpp("test_bessel_j.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 1)

# Compare results with Bessel package
tmp <- Bessel::BesselJ(x, 2, FALSE, 1, 0)
all.equal(tmp, test_bessel_j(x, 2, FALSE, 0))
*/
```

## BesselY (`bessel_y`)

### Description

Computes the Bessel function of the second kind (Neumann function).

### Inputs

- `z`: A (strictly positive) numeric or complex vector of points at which to evaluate the Bessel function.
- `nu`: A numeric value representing the order of the Bessel function.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A numeric or complex vector (depending on the input) with the values of the `bessel_y` function.

### Example

```cpp
// test_bessel_y.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_y(SEXP x, double nu, bool expon_scaled, int verbose) {
  SEXP z = RcppBessel::bessel_y(x, nu, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)
# Source the C++ code
Rcpp::sourceCpp("test_bessel_y.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 0.35)
# skipping zero since bessel_y returns -Inf+0i whereas BesselY return -Inf+NaNi
# Compare results with Bessel package
tmp <- Bessel::BesselY(x, 2, FALSE, 1, 0)
all.equal(tmp, test_bessel_y(x, 2, FALSE, 0))
*/
```



## BesselH (`bessel_h`)

### Description

Computes the Hankel function (Bessel function of the third kind) of the first or second kind.

### Inputs

- `m`: An integer specifying the type of Hankel function. Must be either `1` (for the first kind) or `2` (for the second kind).
- `z`: A numeric or complex vector of points at which to evaluate the Hankel function.
- `nu`: A numeric value representing the order of the Hankel function.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A complex vector with the values of the `bessel_h` function evaluated at the points in `z`.

### Example

```cpp
// test_bessel_h.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_h(int m, SEXP x, double nu, bool expon_scaled, int verbose) {
  SEXP z = RcppBessel::bessel_h(m, x, nu, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)

# Source the C++ code
Rcpp::sourceCpp("test_bessel_h.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 0.35)
# skipping again zero
# Compare results with Bessel package
tmp <- Bessel::BesselH(1, x, 2, FALSE, 1, 0)
all.equal(tmp, test_bessel_h(1, x, 2, FALSE, 0))
*/
```



## AiryA (`airy_a`)

### Description

Computes the Airy function *Ai* for real or complex inputs.

### Inputs

- `z`: A numeric or complex vector of points at which to evaluate the Airy function.
- `deriv`: An integer indicating whether to compute the function itself (`0`) or its first derivative (`1`). Defaults to `0`.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form of the Airy function. Defaults to `FALSE`.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A numeric or complex vector (depending on the input) with the values of the `airy_a` function evaluated at the points in `z`.

### Example

```cpp
// test_airy_a.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_airy_a(SEXP x, int deriv = 0, bool expon_scaled = false, int verbose = 0) {
  SEXP z = RcppBessel::airy_a(x, deriv, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)

# Source the C++ code
Rcpp::sourceCpp("test_airy_a.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 1)

tmp <- Bessel::AiryA(x, 1)
all.equal(tmp, test_airy_a(x, 1, FALSE, 0))
*/
```


## AiryB (`airy_b`)

### Description

Computes the Airy function *Bi* for real or complex inputs.

### Inputs

- `z`: A numeric or complex vector of points at which to evaluate the Airy function.
- `deriv`: An integer indicating whether to compute the function itself (`0`) or its first derivative (`1`). Defaults to `0`.
- `expon_scaled`: A logical value indicating whether to use the exponentially scaled form of the Airy function. Defaults to `FALSE`.
- `verbose`: An integer specifying the verbosity level for error messages.

### Output

- A numeric or complex vector (depending on the input) with the values of the `airy_b` function evaluated at the points in `z`.

### Example

```cpp
// test_airy_b.cpp
#include <Rcpp.h>
#include <RcppBessel.h>
// [[Rcpp::depends(Rcpp, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_airy_b(SEXP x, int deriv = 0, bool expon_scaled = false, int verbose = 0) {
  SEXP z = RcppBessel::airy_b(x, deriv, expon_scaled, verbose);
  return z;
}

/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)

# Source the C++ code
Rcpp::sourceCpp("test_airy_b.cpp")

# Generate a sequence
x <- seq(-10, 10, by = 1)

tmp <- Bessel::AiryB(x, 1)
all.equal(tmp, test_airy_b(x, 1, FALSE, 0))
*/
```

## An Example with RcppArmadillo Conversion

```cpp
#include <RcppArmadillo.h>
#include <RcppBessel.h>

// [[Rcpp::depends(RcppArmadillo, RcppBessel)]]
// [[Rcpp::export]]
SEXP test_bessel_k(const arma::vec& x, double nu, bool expon_scaled = false, int verbose = 0) {
  // Convert arma::vec to Rcpp::NumericVector
  Rcpp::NumericVector rx = Rcpp::wrap(x);
  // Call the BesselK function
  SEXP z = RcppBessel::bessel_k(rx, nu, expon_scaled, verbose);
  // Check if the result is complex
  if (Rcpp::is<Rcpp::ComplexVector>(z)) {
    Rcpp::ComplexVector cz(z);
    arma::cx_vec result = Rcpp::as<arma::cx_vec>(cz);
    return Rcpp::wrap(result);
  } else {
    Rcpp::NumericVector rz(z);
    arma::vec result = Rcpp::as<arma::vec>(rz);
    return Rcpp::wrap(result);
  }
}
/*** R
# Example usage in R
library(Rcpp)
library(RcppBessel)
library(Bessel)
library(RcppArmadillo)

# Source the C++ code
#Rcpp::sourceCpp("test_bessel_k.cpp")

# Generate a sequence
x <- seq(1, 10, by = 1)

# Compare results with Bessel package
tmp <- Bessel::BesselK(x, 2, FALSE, 1, 0)
all.equal(tmp, as.vector(test_bessel_k(x, 2, FALSE, 0)))
*/
```

# Timing Tests

The following code and output provides a quick timing test between the
`Bessel` and `RcppBessel` packages for the `bessel_k` function.

```{r}
library(microbenchmark)
library(Bessel)
library(RcppBessel)
z <- seq(-20, 20, by = 0.001)
b <- microbenchmark(BesselK(z, 2, TRUE), bessel_k(z, 2, TRUE), times = 5L)
print(b)
```



